Open archaeology, open source?

### 2024-02-02 analysis/_03-data_methodology.qmd
Add links to tables

```{r data-oarch}
# TODO: update and replace with local file before submission
read_csv("https://github.com/zackbatist/open-archaeo/raw/2023-08-29/open-archaeo.csv") |>
  mutate(category = na_if(category, "?")) ->
  oarch

@@

```{r tbl-forges}
#| tbl-cap: Software forges used by open archaeology projects
oarch_forges |>
  mutate(p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")) |>
  rename(Host = repo_host, `%` = p) |>
  tt()
```

>>>>

```{r data-oarch}
# TODO: update and replace with local file before submission
read_csv("https://github.com/zackbatist/open-archaeo/raw/master/open-archaeo.csv") |>
  mutate(category = na_if(category, "?")) ->
  oarch

@@

```{r tbl-forges}
#| tbl-cap: Software forges used by open archaeology projects
oarch_forges |>
  mutate(
    repo_host = recode(
      repo_host,
      "GitHub" = "[GitHub](https://github.com/)",
      "Codeberg" = "[Codeberg](https://codeberg.org/)",
      "GitLab" = "[GitLab](https://about.gitlab.com/)",
      "BitBucket" = "[Bitbucket](https://bitbucket.org/product/)",
      "Launchpad" = "[Launchpad](https://launchpad.net/)"
    ),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  rename(Host = repo_host, `%` = p) |>
  tt() |>
  format_tt(1, markdown = TRUE)
```

### 2024-02-02 analysis/_04-open_archaeology.qmd
Add links to tables

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects
tribble(
  ~category,                               ~kind,       ~scope,
  "Packages and libraries",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products",                              "Documents", "Stable outcomes of creative work.",
  "Guides",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(kind, category, scope) |>
  mutate(
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(desc(kind), -n) |>
  select(-kind) |>
  rename_with(tools::toTitleCase) |>
  rename(`%` = p) |>
  tt() |>
  group_tt(list("Software" = 1, "Documents" = 4))
```

@@

```{r tbl-tags}
#| tbl-cap: Themes of open archaeology projects
oarch |>
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) |>
  drop_na(tag) |>
  group_by(tag, category) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(total = sum(n)) |>
  pivot_wider(names_from = category, values_from = n) |>
  mutate(across(everything(), \(x) replace_na(x, 0))) |>
  arrange(-total) |>
  rename(theme = tag) |>
  rename_with(tools::toTitleCase) |>
  tt()
```

@@

```{r tbl-repo-licenses}
#| tbl-cap: Licenses used by open archaeology projects on GitHub
oarch |>
  mutate(license = str_remove(license, "-?\\d\\.\\d$")) |>
  count(license) |>
  mutate(
    license = replace_na(license, "None detected"),
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(-n) |>
  rename_with(tools::toTitleCase) |>
  rename(`%` = p) |>
  tt()
```

>>>>

```{r tbl-categories}
#| tbl-cap: Categories of open archaeology projects
tribble(
  ~category, ~url,                               ~kind,       ~scope,
  "Packages and libraries", "https://open-archaeo.info/categories/packages-and-libraries/",                "Software", "Sets of functions assembled with clear purpose, and made accessible using standards established by an underlying platform.",
  "Standalone software", "https://open-archaeo.info/categories/standalone-software/",                   "Software", "Software that may be operated without needing to first access an underlying platform.",
  "Scripts", "https://open-archaeo.info/categories/scripts/",                               "Software", "Sets of pragmatically assembled mutable functions, often lacking complete documentation or adherence to protocols that would otherwise facilitate secondary use outside their original contexts of creation.",
  "Specifications, protocols and schemas", "https://open-archaeo.info/categories/specifications-protocols-and-schemas/", "Documents", "A formal data structure or framework intended to be used as a model.",
  "Products", "https://open-archaeo.info/categories/products/",                              "Documents", "Stable outcomes of creative work.",
  "Guides", "https://open-archaeo.info/categories/guides/",                                "Documents", "An educational resource or documented protocol meant to instruct readers how to apply relevant tools or techniques.",
  "Lists and datasets", "https://open-archaeo.info/categories/lists-and-datasets/",                    "Documents", "A series of consistently organized observations assembled with purpose."
) |>
  left_join(oarch, by = "category") |>
  count(kind, category, url, scope) |>
  mutate(
    category = glue("[{category}]({url})"),
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(desc(kind), -n) |>
  select(-kind) |>
  rename_with(tools::toTitleCase) |>
  rename(`%` = p) |>
  tt() |>
  group_tt(list("Software" = 1, "Documents" = 4)) |>
  format_tt(1, markdown = TRUE)
```

@@

```{r tbl-tags}
#| tbl-cap: Themes of open archaeology projects
oarch |>
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) |>
  drop_na(tag) |>
  group_by(tag, category) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(total = sum(n)) |>
  pivot_wider(names_from = category, values_from = n) |>
  mutate(
    tag = glue("[{tag}](https://open-archaeo.info/{slugify(tag)})"),
    across(everything(), \(x) replace_na(x, 0))
  ) |>
  arrange(-total) |>
  rename(theme = tag) |>
  rename_with(tools::toTitleCase) |>
  tt()
```

@@

```{r tbl-repo-licenses}
#| tbl-cap: Licenses used by open archaeology projects on GitHub
# TODO: Add links to licenses?
oarch |>
  mutate(license = str_remove(license, "-?\\d\\.\\d$")) |>
  count(license) |>
  mutate(
    license = replace_na(license, "None detected"),
    p = n / sum(n),
    p = paste0(format(p * 100, trim = TRUE, digits = 1), "%")
  ) |>
  arrange(-n) |>
  rename_with(tools::toTitleCase) |>
  rename(`%` = p) |>
  tt()
```

### 2024-02-02 analysis/_05-collaboration.qmd
Add links to tables

```{r fig-lifespan}
#| fig-cap: Lifespan of open archaeology repositories. Each point indicates a commit; excludes repositories with only one commit.
oarch_gh |>
  unnest(gh_commits) |>
  drop_na(datetime) |>
  filter(datetime > ymd("2001-01-02")) |> # Apparent error
  group_by(item_name) |>
  add_tally() |>
  filter(n > 1) |>
  mutate(
    first_commit = min(datetime, na.rm = TRUE),
    mid_commit = median(datetime, na.rm = TRUE),
    last_commit = max(datetime, na.rm = TRUE)
  ) |>
  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
             y = datetime,
             ymin = first_commit,
             ymax = last_commit)) +
  geom_linerange(linewidth = 0.05) +
  geom_point(size = 0.1) +
  scale_y_datetime(date_breaks = "2 years", date_labels = "%Y") +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

>>>>

```{r fig-lifespan}
#| fig-cap: Lifespan of open archaeology repositories. Each point indicates a commit; excludes repositories with only one commit.
fig_lifespan <- oarch_gh |>
  unnest(gh_commits) |>
  drop_na(datetime) |>
  filter(datetime > ymd("2001-01-02")) |> # Apparent error
  group_by(item_name) |>
  add_tally() |>
  filter(n > 1) |>
  mutate(
    first_commit = min(datetime, na.rm = TRUE),
    mid_commit = median(datetime, na.rm = TRUE),
    last_commit = max(datetime, na.rm = TRUE)
  ) |>
  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
             y = datetime,
             ymin = first_commit,
             ymax = last_commit)) +
  geom_linerange(linewidth = 0.05) +
  geom_point(size = 0.1) +
  scale_y_datetime(date_breaks = "2 years", date_labels = "%Y") +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )

girafe(fig_lifespan)
```
